\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\title{Preimage Attack for Reduced Round Keccak}
\author{17111056}
\date{June 2018}

\begin{document}

\maketitle

\section{Preimages on 2 Rounds of the 384-bit Hash function}

\begin{center}
\begin{tikzpicture}[scale=.7]
\draw (0,0) grid (5,5);
    \node[] at (0.5,0.5) {0,0};
    \node[] at (1.5,0.5) {0,1};
    \node[] at (2.5,0.5) {0,2};
    \node[] at (3.5,0.5) {0,3};
    \node[] at (4.5,0.5) {0,4};

    \node[] at (0.5,1.5) {1,0};
    \node[] at (1.5,1.5) {1,1};
    \node[] at (2.5,1.5) {1,2};
    \node[] at (3.5,1.5) {1,3};
    \node[] at (4.5,1.5) {1,4};

    \node[] at (0.5,2.5) {2,0};
    \node[] at (1.5,2.5) {2,1};
    \node[] at (2.5,2.5) {2,2};
    \node[] at (3.5,2.5) {2,3};
    \node[] at (4.5,2.5) {2,4};

    \node[] at (0.5,3.5) {3,0};
    \node[] at (1.5,3.5) {3,1};
    \node[] at (2.5,3.5) {3,2};
    \node[] at (3.5,3.5) {3,3};
    \node[] at (4.5,3.5) {3,4};

    \node[] at (0.5,4.5) {4,0};
    \node[] at (1.5,4.5) {4,1};
    \node[] at (2.5,4.5) {4,2};
    \node[] at (3.5,4.5) {4,3};
    \node[] at (4.5,4.5) {4,4};
    
    \node[anchor=center] at (2.5, -0.5) {Fig 1 : Keccak State};
\end{tikzpicture}
\end{center}

In this section we present a preimage attack for a reduced version with two rounds of Keccak. The preimage can be found in $2^{88}$ time and $2^{87}$ memory. It is not practical, but is an improvement from the present 2 round preimage attack for Keccak-384 which takes $2^{129}$ time [1].

\subsection{Main Scheme}
The Preimage attack on 2 rounds is represented in figure 2.\newline

In Fig 2, each slice represents different states. For the sake of simplicity, we will omit the $\iota$ transformation in the explanation, as it doesn't affect the procedure of the attack, but should be taken into account when implementing the attack (though not practical).

\newpar
We are given a hash value which is 6 lanes in State 4[Fig 2], which represents the output of Keccak permutation. We need to find a message block that produces the same hash value. In Fig2 State 1 the lanes marked with (a0, a1, a2, b0, b1, b2, c0, c1, c2, d0, d1, e0, e1) are the parts of the message that we don't fix until the end of the attack.
The conditions are that : $a2 = a0 \oplus a1$,  $b2 = b0 \oplus b1$, $c2 = c0 \oplus c1$, $d0 = d1 = 0$, $e0 = e1$. Due to these conditions the first $\theta$ operation doesn't affect the state 1 [Fig 2.] From the initial state 1 we can then compute the known lanes in state 2 after $\theta, \rho$ and $\pi$, as well as the positions of the unknown lanes.
\newpar
The change of state from state 1 to state 2 shows the movement of lanes in detail, the number shown in round brackets along with the variable in state 2 is due to $\rho$. Imposing the previous conditions we still have 7*64 degrees of freedom for the message, which is greater than the number of bits of the hash output i.e. 6*64, so we can expect to find atleast one solution.\newline

\begin{center}
\begin{tikzpicture}[scale=1.1]
\draw (0,0) grid (5,5);
    \node[] at (0.5,0.5) {$\mathbf{a_0}$};
    \node[] at (1.5,0.5) {$\mathbf{b_0}$};
    \node[] at (2.5,0.5) {$\mathbf{c_0}$};
    \node[] at (3.5,0.5) {$\mathbf{d_0}$};
    \node[] at (4.5,0.5) {$\mathbf{e_0}$};

    \node[] at (0.5,1.5) {$\mathbf{a_2}$};
    \node[] at (1.5,1.5) {$\mathbf{b_2}$};
    \node[] at (2.5,1.5) {$\mathbf{c_1}$};
    \node[] at (3.5,1.5) {$\mathbf{d_1}$};
    \node[] at (4.5,1.5) {$\mathbf{e_1}$};

    \node[] at (0.5,2.5) {$\mathbf{a_1}$};
    \node[] at (1.5,2.5) {$\mathbf{b_1}$};
    \node[] at (2.5,2.5) {$\mathbf{c_2}$};
    \node[] at (3.5,2.5) {0};
    \node[] at (4.5,2.5) {0};

    \node[] at (0.5,3.5) {0};
    \node[] at (1.5,3.5) {0};
    \node[] at (2.5,3.5) {0};
    \node[] at (3.5,3.5) {0};
    \node[] at (4.5,3.5) {0};

    \node[] at (0.5,4.5) {0};
    \node[] at (1.5,4.5) {0};
    \node[] at (2.5,4.5) {0};
    \node[] at (3.5,4.5) {0};
    \node[] at (4.5,4.5) {0};

\node[anchor=center] at (2.5, -0.5) {State 1};

\draw (6,0) grid (11,5);
    \node[] at (6.5,0.5) {$a_0$(0)};
    \node[] at (7.5,0.5) {$b_2$(44)};
    \node[] at (8.5,0.5) {$c_2$(43)};
    \node[] at (9.5,0.5) {0};
    \node[] at (10.5,0.5) {0};

    \node[] at (6.5,1.5) {$d_0$(28)};
    \node[] at (7.5,1.5) {$e_1$(20)};
    \node[] at (8.5,1.5) {$a_1$(3)};
    \node[] at (9.5,1.5) {0};
    \node[] at (10.5,1.5) {0};

    \node[] at (6.5,2.5) {$b_0$(1)};
    \node[] at (7.5,2.5) {$c_1$(6)};
    \node[] at (8.5,2.5) {0};
    \node[] at (9.5,2.5) {0};
    \node[] at (10.5,2.5) {0};

    \node[] at (6.5,3.5) {$e_0$(27)};
    \node[] at (7.5,3.5) {$a_2$(36)};
    \node[] at (8.5,3.5) {$b_1$(10)};
    \node[] at (9.5,3.5) {0};
    \node[] at (10.5,3.5) {0};

    \node[] at (6.5,4.5) {$c_0$(62)};
    \node[] at (7.5,4.5) {$d_1$(55)};
    \node[] at (8.5,4.5) {0};
    \node[] at (9.5,4.5) {0};
    \node[] at (10.5,4.5) {0};

\node[anchor=center] at (8.5, -0.5) {State 2};
    \draw[thick,->] (5, 2.5) -- node [above] {$\theta, \rho, \pi$} (6,2.5);
    \draw[thick,->] (9, -0.8) -- node [left] {$\chi, \iota, \theta$} (9, -2);
\end{tikzpicture}
\end{center}
% \[
% \]
\begin{center}
\begin{tikzpicture}[scale=1]
\draw (2,0) grid (7,5);
    \node[] at (2.5,0.5) {$h_0$};
    \node[] at (3.5,0.5) {$h_1$};
    \node[] at (4.5,0.5) {$h_2$};
    \node[] at (5.5,0.5) {$h_3$};
    \node[] at (6.5,0.5) {$h_4$};

    \node[] at (2.5,1.5) {$h_5$};
    \node[] at (3.5,1.5) { };
    \node[] at (4.5,1.5) { };
    \node[] at (5.5,1.5) { };
    \node[] at (6.5,1.5) { };

    \node[] at (2.5,2.5) { };
    \node[] at (3.5,2.5) { };
    \node[] at (4.5,2.5) { };
    \node[] at (5.5,2.5) { };
    \node[] at (6.5,2.5) { };

    \node[] at (2.5,3.5) { };
    \node[] at (3.5,3.5) { };
    \node[] at (4.5,3.5) { };
    \node[] at (5.5,3.5) { };
    \node[] at (6.5,3.5) { };

    \node[] at (2.5,4.5) { };
    \node[] at (3.5,4.5) { };
    \node[] at (4.5,4.5) { };
    \node[] at (5.5,4.5) { };
    \node[] at (6.5,4.5) { };

\draw (10,0) grid (15,5);
    \node[] at (10.5,0.5) {h0'};
    \node[] at (11.5,0.5) { };
    \node[] at (12.5,0.5) { };
    \node[] at (13.5,0.5) {h5'};
    \node[] at (14.5,0.5) { };

    \node[] at (10.5,1.5) { };
    \node[] at (11.5,1.5) {h1'};
    \node[] at (12.5,1.5) { };
    \node[] at (13.5,1.5) { };
    \node[] at (14.5,1.5) {1};

    \node[] at (10.5,2.5) { };
    \node[] at (11.5,2.5) { };
    \node[] at (12.5,2.5) {h2'};
    \node[] at (13.5,2.5) { };
    \node[] at (14.5,2.5) { };

    \node[] at (10.5,3.5) { };
    \node[] at (11.5,3.5) { };
    \node[] at (12.5,3.5) { };
    \node[] at (13.5,3.5) {h3'};
    \node[] at (14.5,3.5) { };

    \node[] at (10.5,4.5) { };
    \node[] at (11.5,4.5) { };
    \node[] at (12.5,4.5) { };
    \node[] at (13.5,4.5) { };
    \node[] at (14.5,4.5) {h4'};
    
    \node[anchor=center] at (10.5, -0.5) {State 3};
    \node[anchor=center] at (4.5, -0.5) {State 4};
     \draw[thick,->] (7, 2.5) -- node [above] {$\iota^{-1}, \chi^{-1}, \pi^{-1}, \rho^{-1}$} (10,2.5);
     \node[anchor=center] at (8.5, -1) {Fig 2: Diagram of the 2-round preimage attack. Each square represents a 64 bit lane.};
\end{tikzpicture}
\end{center}    
\newpage
\newpar
In the backward direction, we can invert from State 4, the known 6 lanes in final state with $\chi^{-1}$. The first five lanes can be directly inverted using $\chi^{-1}$, for the remaining sixth lane, we can fix the seventh lane of State 5 to \textbf{1} for inverting sixth lane. As shown in Fig 3. Then, we can apply the inverse of $\pi$ and of $\rho$ and obtain the values and positions of the 6 known lanes in State 3.

\begin{center}
\begin{tikzpicture}[scale=1]
\draw (2,0) grid (7,5);
    \node[] at (2.5,0.5) {$h_0$};
    \node[] at (3.5,0.5) {$h_1$};
    \node[] at (4.5,0.5) {$h_2$};
    \node[] at (5.5,0.5) {$h_3$};
    \node[] at (6.5,0.5) {$h_4$};

    \node[] at (2.5,1.5) {$h_5$};
    \node[] at (3.5,1.5) { };
    \node[] at (4.5,1.5) { };
    \node[] at (5.5,1.5) { };
    \node[] at (6.5,1.5) { };

    \node[] at (2.5,2.5) { };
    \node[] at (3.5,2.5) { };
    \node[] at (4.5,2.5) { };
    \node[] at (5.5,2.5) { };
    \node[] at (6.5,2.5) { };

    \node[] at (2.5,3.5) { };
    \node[] at (3.5,3.5) { };
    \node[] at (4.5,3.5) { };
    \node[] at (5.5,3.5) { };
    \node[] at (6.5,3.5) { };

    \node[] at (2.5,4.5) { };
    \node[] at (3.5,4.5) { };
    \node[] at (4.5,4.5) { };
    \node[] at (5.5,4.5) { };
    \node[] at (6.5,4.5) { };

\draw (10,0) grid (15,5);
    \node[] at (10.5,0.5) {h0'};
    \node[] at (11.5,0.5) {h1'};
    \node[] at (12.5,0.5) {h2'};
    \node[] at (13.5,0.5) {h3'};
    \node[] at (14.5,0.5) {h4'};

    \node[] at (10.5,1.5) {h5'};
    \node[] at (11.5,1.5) {\textbf{1}};
    \node[] at (12.5,1.5) { };
    \node[] at (13.5,1.5) { };
    \node[] at (14.5,1.5) { };

    \node[] at (10.5,2.5) { };
    \node[] at (11.5,2.5) { };
    \node[] at (12.5,2.5) { };
    \node[] at (13.5,2.5) { };
    \node[] at (14.5,2.5) { };

    \node[] at (10.5,3.5) { };
    \node[] at (11.5,3.5) { };
    \node[] at (12.5,3.5) { };
    \node[] at (13.5,3.5) { };
    \node[] at (14.5,3.5) { };

    \node[] at (10.5,4.5) { };
    \node[] at (11.5,4.5) { };
    \node[] at (12.5,4.5) { };
    \node[] at (13.5,4.5) { };
    \node[] at (14.5,4.5) { };
    
    \node[anchor=center] at (12.5, -0.5) {State 5};
    \node[anchor=center] at (4.5, -0.5) {State 4};
     \draw[thick,->] (7.7, 2.5) -- node [above] {$\chi^{-1}$} (9.7,2.5);
     \node[anchor=center] at (8.5, -1.2) {Fig 3: Inversion of $\chi^{-1}$, with the condition h5 = h5'};
\end{tikzpicture}
\end{center}    

Note : Due to $\chi$, h5 = h5' $\oplus$ ( $\bar 1$ $\cdot$ h6') $\to$ h5 = h5' .

\newpar
Then, the issue is to find the values of the eleven 64-bit words (a0, a1, a2, b0, b1, b2, c0, c1, c2, e0, e1) in state 2 that make possible the transition by the operations $\chi$ and $\theta$ from the state 2, that we have obtained computing forward, to the state 3, that we have obtained computing backward. For this we start by finding the bits that verify the relations for few slices.  The idea is same as in [2]. The idea is, to consider first groups of 3 slices where we guess all the involved bits of a0,...,e1 and next we can do a sieving by just keeping the guesses ones that produce by $\chi$ and $\theta$ the values of the 7x2 = 14 known bits from State 3 in the middle and last slices of the group of three slices.This is possible as for computing the output of $\theta$ in a specific slice, we need to know this same slice and the previous one in the input state.

\subsection{Finding Partial Solutions}

\newpar
Partial solutions for 3 slices : We start by finding partial solutions for groups of three slices. In three slices there are 3*11 bit-variables from a0,..e1. If we consider the conditions then also there are no repetitions, so we have o guess in total 33 variables. For each guess we check if the 14 bits already fixed computing backwards on the two last slices collide, which happens with probability $2^{-14}$. So for each group of three slices that we find partial solutions for, we get $2^{33 - 14} = 2^{19}$ solutions. This will be repeated for 16 consecutive groups of 3 slices, leaving 16 out. This has time and memory complexity of $16 * 2^{19} = 2^{23}$.

\newpar
Partial solutions for 6 slices : For 6 slcies we merge two consecutive groups of three slices. As per conditions there is no repetition. So we have $2^{19*2}$ possibilities. When we merge two groups, we obtain values of 7 bits of the first slice of the second group. So we get total $2^{19*2 - 7 } = 2^{31}$ solutions. There are 8 groups of 6 slices. So this step has time and memory complexity of $8*2^{31} = 2^{34}$.

\newpar
Partial solutions for 12 slices : Similarly, we merge here each two consecutive groups of 6 slices. In this case the number of repeated bits in merge is 5 (from e0,e1), so total number of solutions that we obtain is $2^{31*2 - 5 - 7 } = 2^{50}$. There are 4 groups of 12 slices, so it has time and memory complexity of $4*2^{50} = 2^{52}$.

\newpar
Partial solutions for 24 slices : Similarly, we merge here each two consecutive groups of 24 slices. In this case the number of repeated bits in merge is 7 (from e0,e1), so total number of solutions that we obtain is $2^{50*2 - 7 - 7 } = 2^{86}$. There are 2 groups of 24 slices, so it has time and memory complexity of $2*2^{86} = 2^{87}$.

\newpar
Partial solutions for 48 slices : Finally here, we merge here two groups of 24 slices. In this case the number of repeated bits in merge is 29 (from a0, a1, a2), 23 (from b0, b1, b2), 24 (from c0, c1, c2) and 7 (from e0,e1), so total number of solutions that we obtain is $2^{86*2 - (29 + 23 + 24 + 7) - 7 } = 2^{82}$. There are 2 groups of 24 slices, so it has time and memory complexity of $2^{82}$.

\newpar
Partial solutions for 16 slices : For finding solutions for 16 slices we first find solutions for the 12 rightmost slices the same way as before, and obtaining $2^{50}$ partial solutions. 
Next, we obtain solutions for the remaining 4 slices, we have 44 variables and none of them are repeated. Also we have to satisfy last 3 slices out of the 4. This gives $2^{44 - 7*3} = 2^{23}$ solutions.
Now, we can merge the group of 12 slices and 4 slices to obtain partial solutions for 16 slices. Among the 12 and 4 slices groups there are 4 repetitions(from e0, e1) and additionally we obtain values of 7 bits of the first slice of the 4 slice group which gives us $2^{50 + 23 - 4 -7} = 2^{62}$ solutions.

Now, we have to merge the solutions for 48 slices and 16 slices. They have in common 35 ( from a0, a1, a2), 41 (from b0, b1, b2), 40 (from c0, c1 c2) and 14 (from e0, e1). Additionally we obtain values of 7 bits of the first slice of the 48 and 16 slice groups which gives us $2^{82 + 62 - (35+ 41 + 40 + 14) - 2*7} = 2^{0} = 1$ solution. This step has time complexity $ = 2^{82}$. 

Total time complexity of the attack is given by $2^{33} + 2^{34} + 2^{52} + 2^{87} + 2^{63} + 2^{82} = 2^{88}$ and memory $2^{87}$. 

This confirms that there exists a set of values for the variables such that the preimage can be obtained from the hash value for keccak-384.

\section{References :}

1. Jian Guo and Meicheng Liu and Ling Song, Linear Structures: Applications to Cryptanalysis of Round-Reduced Keccak.\newline

\end{document}
